Efficient Thresholded Correlation using Truncated Singular Value Decomposition

### 2016-03-02 threshold/draft.Rnw
Remove jss format

\documentclass[article]{jss}
%% need no \usepackage{Sweave.sty}

\usepackage{amsmath}
\usepackage{hyperref}

@@

\newcommand{\qed}{\nobreak \ifvmode \relax \else
      \ifdim\lastskip<1.5em \hskip-\lastskip
      \hskip1.5em plus0em minus0.5em \fi \nobreak
      \vrule height0.75em width0.5em depth0.25em\fi}

\title{Efficient Thresholded Correlation using Truncated Singular Value Decomposition}
\author{
James Baglama\\
University of Rhode Island
\And
Michael Kane\\
Yale University
\AND
Bryan Lewis\\
Paradigm4, Inc.\\
\And
Alex Poliakov\\
Paradigm4, Inc.
}
\Plainauthor{James Baglama, Michael Kane, Bryan Lewis, Alex Poliakov}
\Plaintitle{Efficient Thresholded Correlation using Truncated Singular Value Decomposition}
\Shorttitle{Efficient Thresholded Correlation}

\Abstract{
Efficiently computing a subset of a correlation matrix consisting of values
above a specified threshold is important to many practical applications.
Real-world problems in genomics, machine learning, finance other applications
can produce correlation matrices too large to explicitly form and tractably
compute. Often, only values corresponding to highly-correlated vectors are of
interest, and those values typically make up a small fraction of the overall
correlation matrix. We present a method based on the singular value
decomposition (SVD) and its relationship to the data covariance structure that
can efficiently compute thresholded subsets of very large correlation matrices.
}
\Keywords{thresholded correlation, SVD, \proglang{R}}
\Plainkeywords{thresholded correlation, SVD, R}
\Address{
James Baglama, Chair of Mathematics\\
Department of Mathematics\\
University of Rhode Island\\
Kingston, Rhode Island 02881\\
E-mail: \email{jbaglama@math.uri.edu}\\
URL: \url{http://www.math.uri.edu/~jbaglama}\\
\\
Michael Kane\\
Yale University\\
300 Geroge Street\\
New Haven, Connecticut 06511\\
E-mail: \email{michael.kane@yale.edu}\\
\\
Bryan Lewis\\
Paradigm4, Inc.\\
281 Winter Street, Suite 360\\
Waltham, MA 02451\\
E-mail: \email{blewis@illposed.net}\\
URL: \url{http://illposed.net}\\
\\
Alex Poliakov\\
Paradigm4, Inc.\\
281 Winter Street, Suite 360\\
Waltham, MA 02451\\
E-mail: \email{apoliakov@paradigm4.com}\\
}

\begin{document}

@@

Anyone can quickly run the example to experiment with the algorithm and
compare its output with brute force results. The example uses the {\tt
EisenYeast} data from the {\tt biclust} \proglang{R} package \citep{biclust}.
The data consist of 80 experiments involving the
Saccharomyces Cerevisiae yeast measuring gene expression data across 6,221
Affymetrix probes, represented as an $80\times 6221$ real-valued matrix.  The
example finds all pairs of columns (gene expression measurements) that meet or exceed a
Pearson correlation value of 0.95. The choice of projection dimension $p$ affects
the efficiency of the algorithm. For $p=2$ the number of candidate vector pairs
generated in step 4 of Algorithm \ref{prune} is over 3 million. When $p$ is increased
to $10$, the number of candidate vector pairs drops to less than 13,000.
The reference \proglang{R} implementation includes options for
tuning $p$ and relies on the restart ability of the IRLBA to pick up where it left off and
inexpensively extend an existing truncated SVD. The following example uses $p=10$.
\begin{small}
<<eval=TRUE,echo=TRUE>>=
threshold = 0.95

>>>>

\documentclass{article}

\usepackage{Sweave}
\usepackage{amsmath}
\usepackage{hyperref}
\usepackage{cite}
\usepackage{authblk}
\usepackage{graphicx}
\let\proglang=\textsf

@@

\title{Efficient Thresholded Correlation using Truncated Singular Value Decomposition}
%\author{
%James Baglama\\
%University of Rhode Island
%\And
%Michael Kane\\
%Yale University
%\AND
%Bryan Lewis\\
%Paradigm4, Inc.\\
%\And
%Alex Poliakov\\
%Paradigm4, Inc.
%}
%\Plainauthor{James Baglama, Michael Kane, Bryan Lewis, Alex Poliakov}
%\Plaintitle{Efficient Thresholded Correlation using Truncated Singular Value Decomposition}
%\Shorttitle{Efficient Thresholded Correlation}
\author[1]{James Baglama \thanks{jbaglama@math.uri.edu}}
\author[2]{Michael Kane \thanks{michael.kane@yale.edu}}
\author[3]{Bryan Lewis \thanks{blewis@illposed.net}}
\author[3]{Alex Poliakov \thanks{apoliakov@paradigm4.com}}
\affil[1]{
University of Rhode Island
}
\affil[2]{
Yale University
}
\affil[3]{
Paradigm4, Inc.
}
\renewcommand\Authands{ and }

%\Abstract{
\begin{document}
\maketitle
\begin{abstract}
Efficiently computing a subset of a correlation matrix consisting of values
above a specified threshold is important to many practical applications.
Real-world problems in genomics, machine learning, finance other applications
can produce correlation matrices too large to explicitly form and tractably
compute. Often, only values corresponding to highly-correlated vectors are of
interest, and those values typically make up a small fraction of the overall
correlation matrix. We present a method based on the singular value
decomposition (SVD) and its relationship to the data covariance structure that
can efficiently compute thresholded subsets of very large correlation matrices.
\end{abstract}
%\Keywords{thresholded correlation, SVD, \proglang{R}}
%\Plainkeywords{thresholded correlation, SVD, R}
%\Address{
%James Baglama, Chair of Mathematics\\
%Department of Mathematics\\
%University of Rhode Island\\
%Kingston, Rhode Island 02881\\
%E-mail: \email{jbaglama@math.uri.edu}\\
%URL: \url{http://www.math.uri.edu/~jbaglama}\\
%\\
%Michael Kane\\
%Yale University\\
%300 Geroge Street\\
%New Haven, Connecticut 06511\\
%E-mail: \email{michael.kane@yale.edu}\\
%\\
%Bryan Lewis\\
%Paradigm4, Inc.\\
%281 Winter Street, Suite 360\\
%Waltham, MA 02451\\
%E-mail: \email{blewis@illposed.net}\\
%URL: \url{http://illposed.net}\\
%\\
%Alex Poliakov\\
%Paradigm4, Inc.\\
%281 Winter Street, Suite 360\\
%Waltham, MA 02451\\
%E-mail: \email{apoliakov@paradigm4.com}\\
%}

@@

Anyone can quickly run the example to experiment with the algorithm and
compare its output with brute force results. The example uses the {\tt
EisenYeast} data from the {\tt biclust} \proglang{R} package \cite{biclust}.
The data consist of 80 experiments involving the
Saccharomyces Cerevisiae yeast measuring gene expression data across 6,221
Affymetrix probes, represented as an $80\times 6221$ real-valued matrix.  The
example finds all pairs of columns (gene expression measurements) that meet or exceed a
Pearson correlation value of 0.95. The choice of projection dimension $p$ affects
the efficiency of the algorithm. For $p=2$ the number of candidate vector pairs
generated in step 4 of Algorithm \ref{prune} is over 3 million. When $p$ is increased
to $10$, the number of candidate vector pairs drops to less than 13,000.
The reference \proglang{R} implementation includes options for
tuning $p$ and relies on the restart ability of the IRLBA to pick up where it left off and
inexpensively extend an existing truncated SVD. The following example uses $p=10$.
\begin{small}
<<eval=TRUE,echo=TRUE>>=
threshold = 0.95
